---
title: "exploratory-konza"
author: "Sarah Supp"
date: "6/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

library(tseries)
library(betapart)
library(vegan)

set.seed(123)
```

bring in the Konza Praire small mammal burn study data from 1994-2011 (19 years total). 
Burns at varying intevals, can compare periodic disturbance effects from no, infrequent, frequent
spB; not burned?
FB; Fall Burn; burned only in wet years
SuB; Burned biennially (every 2 years)
WA; Winter Burn; unburned for 3 consecutive years, then burned for 3 consecutive years
FA; Fall Burn; 4 year burn treatment
WB; Winter Burn; 4 year burn treatment
004b; remove, not sure?
```{r}
dat <- read_csv("CSM041.csv")
surv <- read_csv("CSM042.csv")
```
Make some basic exploratory plots and summaries
***warning*** I have not yet investigated or controlled for sampling variation among years, and will want to correct for that before putting anything into a document, proposal, etc. Just an initial start to getting code down that will work as proof of concept.

Plot the counts of species (lump spring and fall samples) across years and plot
```{r}
#remove treatment/watershed 004b because I don't know what it is
#remove dec because it's not sampled for all treatments in all years
surv <- surv %>%
  filter(Watershed != "004b") %>%
  filter(Season != "dec") %>%
  filter(Recyear != 1993) %>%
  filter(Species != "x" & Species !="p?")

#summing across spring and autumn samples, because lumping annually makes sense here
surv %>%
  group_by(Watershed, Recyear) %>%
  summarize(spcount = n_distinct(Species)) %>%
  ggplot(aes(Recyear, spcount, group=Watershed)) + geom_line(aes(col=Watershed)) + facet_wrap(~Watershed)
```
Calculate species persistence as proportion of years present for each treatment
```{r}
surv %>%
  group_by(Watershed, Species) %>%
  summarize(num_years=n_distinct(Recyear)) %>%
  ggplot(aes(Species, num_years)) + geom_point(aes(col=Watershed)) + facet_wrap(~Watershed)

persistence <- surv %>%
  filter(Recyear != 1993) %>%
  group_by(Watershed, Species) %>%
  summarize(num_years = n_distinct(Recyear), 
            prop_years = round(num_years/18,3))

ggplot(persistence, aes(Species, prop_years)) + geom_point(aes(col=Watershed)) + facet_wrap(~Watershed)

```

Try Gotelli and Faye's code out for incidence patterns
First, the functions
```{r}
#Classification scheme
getTrends<-function(x) {
  
  time <- rep("late",length(x))
  time[1:(round(length(x)/2))] <- "early"
  
  # tabulate f
  z_x <- table(x,time)
  
  tslen<-length(x)
  tssum<-sum(x)
  l<-c()
  
  for(k in 1:(tslen-1)){
    j<-abs(x[k+1]-x[k])
    l<-c(l, j)
    v<-sum(l)
  }
  
  bsline<-x[1]
  
  if(tslen>tssum) {
    
    rownames(z_x) <- c("absent","present")
    # get p value
    p_val <- chisq.test(z_x)$p.val
    
    # get early and late fractions
    f_early <- z_x["present","early"]/sum(z_x[,1])
    f_late <- z_x["present","late"]/sum(z_x[,2])
   
    if((f_early > f_late) & (p_val <= 0.05)) trend<--1
    if((f_early < f_late) & (p_val <= 0.05)) trend<-1
    if(p_val > 0.05) trend<-0
    runsPV<-runs.test(as.factor(x),alternative="less")
    runsTestPV<-runsPV$p.value
    if(runsTestPV<0.05) trendPlus<-10
    if(runsTestPV>0.05)trendPlus<-5
  }
  if(tslen==tssum) {
    p_val <- NA
    f_early <- NA
    f_late <- NA
    runsTestPV<-NA
    trend<-NA
    trendPlus<-NA
  }
  
  statSumm<-list("trendPlus"=trendPlus, "chiPval"=p_val, "runsPval"=runsTestPV, 
                 "trend"=trend, "bsline"=bsline)
  return(statSumm)
} 


#Calculating compositional dissimilarity
getBeta<-function(df) {
  
  yr<-unique(df[, 1]) 
  df<-df[,-1]
  bindf<-df 
  bindf[bindf>0]<-1
  getBas<-beta.pair(df, index.family="jac")$beta.jac
  getChao<-vegdist(bindf, method="chao")
  
  basJac<-c(1, getBas[1:(nrow(spec_df)-1)])
  chaoJac<-c(1, getChao[1:(nrow(spec_df2)-1)])
  
  df_out<-data.frame(Year=yr, J=basJac, chaoJac=chaoJac)
 
  return(df_out)
} 

#Randomisation functions
## where t1 is the long form time series and p1 is the subset of species with the classification to be randomised

### this first function takes account only of presence and absence

getRandomisation<-function(t1, p1) {
  
  df<-data.frame()
  nameRow<-c()
  
  m1<-as.data.frame(pivot_wider(t1, names_from=Year, values_from=Abundance))
  m1[is.na(m1)]<-0
  nameCol<-unique(t1$Year)
  
  for(i in 1:length(m1$Species)) {
    x<-m1[i,]
    sp<-as.vector(m1[i,1])
    dird<-as.vector(unique(p1$Species))
    if(sp %in% dird) {
      nameRow<-c(nameRow, sp)
      x<-x[-1]
      x[x>0]<-1
      y<-as.data.frame(t(x))[,1]
      v<-sample(y[-1])
      z<-c(y[1], v)
      df<-rbind(df, z)
    }
    if(!sp %in% dird) {
      nameRow<-c(nameRow, sp)
      x<-x[-1]
      x[x>0]<-1
      y<-as.data.frame(t(x))[,1]
      z<-y
      df<-rbind(df, z)
    }
  }
  
  colnames(df)<-nameCol
  rownames(df)<-nameRow
 
  nameCR<-c("Species", nameCol)
  names(df)<-nameCR
  df2<-as.data.frame(t(df))
  df3<-df2[-1,]
  names(df3)
  names(df3)<-nameRow
  
  return(df3)
}

### function to replace sample to shuffle weighted by abundance

abun_shuffle <- function() {
  
  occur_shuf <- rep(0,length(occur)) # set up an empty vector
  index <- seq_along(occur) # set up vector of indices
  presence <- sample(x=index,size=sum(occur),prob=abun) # select index values 
  occur_shuf[presence] <- 1 #assign occurrences
  return(occur_shuf)
} # end of abun_shuffle
# -------------------------------------


## revised randomisation function which now includes the vector of abundances for a time series

getAbundRandomisation<-function(t1, p1, rf1) {
  
  df<-data.frame()
  nameRow<-c()
  
  m1<-as.data.frame(pivot_wider(t1, names_from=Year, values_from=Abundance))
  m1[is.na(m1)]<-0
  nameCol<-unique(t1$Year)
  
  for(i in 1:length(m1$Species)) {
    x<-m1[i,]
    sp<-as.vector(m1[i,1])
    dird<-as.vector(unique(p1$Species))
    if(sp %in% dird) {
      nameRow<-c(nameRow, sp)
      x<-x[-1]
      x[x>0]<-1
      y<-as.data.frame(t(x))[,1]
      v<-abun_shuffle(y[-1], rf1[-1])
      z<-c(y[1], v)
      df<-rbind(df, z)
    }
    if(!sp %in% dird) {
      nameRow<-c(nameRow, sp)
      x<-x[-1]
      x[x>0]<-1
      y<-as.data.frame(t(x))[,1]
      z<-y
      df<-rbind(df, z)
    }
  }
  
  colnames(df)<-nameCol
  rownames(df)<-nameRow
  
  nameCR<-c("Species", nameCol)
  names(df)<-nameCR
  df2<-as.data.frame(t(df))
  df3<-df2[-1,]
  names(df3)
  names(df3)<-nameRow
  
  return(df3)
}

#Other functions
getSummStats<-function(x, y, namesCl) {
  x<-na.omit(x)
  y<-na.omit(y)
  names(x)<-c("nx")
  names(y)<-c("ny")
  meanCl<-mean(x$nx)
  medianCl<-median(x$nx)
  sd<-sd(x$nx)
  noRF<-sum(y$ny)
  confCl<-round(quantile(x$nx, probs=c(0.025,0.975)), 4)
  v<-c("class"=namesCl, "mean"=meanCl, "median"=medianCl, "sd"=sd, "number of populations"=noRF, "confid levels"=confCl)
  return(v)
}
```

Then, construct species occurrence matrices, where row is species, column is year, and presence=1 and absence=0
#fixme: I think this is starting to work correctly, but double check.
Increasing: chiPval <0.05 and trend = 1
Decreasing: chiPval <0.05 and trend = -1
Recurrent: runsPval <0.05 and trendPlus = 10
```{r}
# This is how we run it for each treatment. Need to compile the results better for summarizing and plotting

#create an empty dataframe to store the results
res <- data.frame(species = "sp", treatment = "trmt", prop=0, traj="traj", trendPlus=0, chiPval=0, runsPval=0, trend=0, bsline=0)

incid <- data.frame(species=NA, year=NA, incidence=NA, treatment=NA)

#identify unique treatments
treatment <- unique(surv$Watershed)

for(t in treatment){
  trmt=t

  trmt_matrix <- surv %>%
  filter(Watershed == t) %>% #0spb, 0sub
  group_by(Recyear, Species)  %>%
  summarize(n=n()) %>%
  mutate_at(vars(n), ~replace(., . > 1, 1)) %>%
  mutate(row=row_number()) %>%
  spread(Recyear, row, fill=0) %>%
  mutate(across(3:"2011", ~replace(., . >1, 1)))
  
  long <- trmt_matrix %>%
    select(-n) %>%
    pivot_longer(!Species, names_to = "year", values_to = "incidence") %>%
    mutate(treatment=t) %>%
    rename(species=Species)
  incid <- rbind(incid, long)
  
for(i in 1:nrow(trmt_matrix)) {
  sp = trmt_matrix[i, 1][[1]]
  spdat <- as.numeric(as.matrix(trmt_matrix[i,c(3:ncol(trmt_matrix))]))
  #get proportion of years present
  prop <- sum(spdat)/length(spdat)
  #could write a series of ifelse statements to look at output and label by category, then save into a table with relevant numeric result labeled by each species
  if(sum(spdat) == length(spdat)) {
    traj <- "no change"
    res <- rbind(res, data.frame(species=sp, treatment=trmt, prop=prop, traj=traj,
                                 trendPlus=NA, chiPval=NA, runsPval=NA, trend=NA, bsline=NA))
    next
  }
  else{
    vals <- getTrends(spdat)
    if(vals$chiPval < 0.05) {
      if(vals$trend == 1) { traj = "increasing"
      }
      else { traj = "decreasing"
      }
      res <- rbind(res, data.frame(species=sp, treatment=trmt, prop=prop, traj=traj, 
                                   trendPlus=vals$trendPlus, chiPval=vals$chiPval, runsPval=vals$runsPval,
                                   trend=vals$trend, bsline=vals$bsline))
    }
    else if (vals$runsPval <0.05) { traj = "recurrent"
          res <- rbind(res, data.frame(species=sp, treatment=trmt, prop=prop, traj=traj, trendPlus=vals$trendPlus,
                                       chiPval=vals$chiPval, runsPval=vals$runsPval, trend=vals$trend,
                                       bsline=vals$bsline))
    }
    else { traj = "random"
          res <- rbind(res, data.frame(species=sp, treatment=trmt, prop=prop, traj=traj, 
                                       trendPlus=vals$trendPlus, chiPval=vals$chiPval, runsPval=vals$runsPval,
                                       trend=vals$trend, bsline=vals$bsline))
    }
  }
}
}
res <- res[-1,]
incid <- incid[-1,]
#fa_matrix: 8 spp; no increasing or decreasing species, 2 recurrent species (all others random or no change)
#fb_matrix: 8 spp; 1 increasing, 1 recurrent, no decreasing
#wa_matrix: 10 spp; no increasing or decreasing species, 2 recurrent
#wb_matrix: 10 spp; all random or no change
#spb_matrix: 13 spp; 1 increasing
#sub_matrix: 12 spp; 1 recurrent
```


OK now that we have the results, let's compare the proportions, and trajectories
```{r}
res %>%
  group_by(treatment, traj) %>%
  summarize(count=n()) %>%
  ggplot(aes(traj, count)) + geom_bar(stat="identity", aes(fill=treatment), position="dodge")
```

merging datasets for a bigger plot
spB; not burned?
FB; Fall Burn; burned only in wet years
SuB; Burned biennially (every 2 years)
WA; Winter Burn; unburned for 3 consecutive years, then burned for 3 consecutive years
FA; Fall Burn; 4 year burn treatment
WB; Winter Burn; 4 year burn treatment
```{r}
#merge the results with the incidence dataframe for later plotting
test <- inner_join(incid, res) %>%
  mutate(treatment = replace(treatment, treatment=="0spb", "0 burns"),
         treatment = replace(treatment, treatment=="00fa", "4-yr fall burn"),
         treatment = replace(treatment, treatment=="00fb", "wet year fall burn"),
         treatment = replace(treatment, treatment=="00wa", "3-yr con winter burn"),
         treatment = replace(treatment, treatment=="00wb", "4-yr winter burn"),
         treatment = replace(treatment, treatment=="0sub", "2-yr burn"))


ggplot(test, aes(prop, year, fill=as.factor(incidence))) + 
  geom_tile(alpha=0.5, aes(col=species)) + 
  scale_fill_manual(values=c("white","blue")) +
  facet_grid(treatment~traj) + # compare with wrapping on treatment
  theme_bw() + 
  theme(legend.position = "none")
```


compare species trends
```{r}
unique(res$species)
res %>%
  filter(species == "rmont")

#PL switches: recurrent on 00fa, nochange on 0sub, random all others
#RM switches: increasing on 00fb, random all others
#ST switches: recurrent on 00fb and 00wa, random all others
#CH switches: recurrent on 00wa, random others (not present on all)
#NF switches: increasing on 0spb, random others (not present on all)
```

compare proportions for each category
```{r}
res %>%
  group_by(treatment) %>%
  summarize(spcount=n_distinct(species), 
            trajcount=n_distinct(traj)) %>%
  ggplot(aes(spcount, trajcount)) + geom_point(aes(col=treatment)) + stat_smooth(method="lm")

proptraj <- res %>%
  group_by(treatment) %>%
  mutate(spcount=n_distinct(species), 
            trajcount=n_distinct(traj)) %>%
  group_by(treatment, traj) %>%
  summarize(prop_traj=n()/spcount,
            prop_avg=mean(prop))

res <- res %>%
    mutate(treatment = replace(treatment, treatment=="0spb", "0 burns"),
         treatment = replace(treatment, treatment=="00fa", "4-yr fall burn"),
         treatment = replace(treatment, treatment=="00fb", "wet year fall burn"),
         treatment = replace(treatment, treatment=="00wa", "3-yr con winter burn"),
         treatment = replace(treatment, treatment=="00wb", "4-yr winter burn"),
         treatment = replace(treatment, treatment=="0sub", "2-yr burn"))

ggplot(proptraj, aes(treatment, prop_traj)) + geom_bar(aes(fill=traj), stat="identity", position="dodge", col="white")

ggplot(proptraj, aes(treatment, prop_avg)) + geom_bar(aes(fill=traj), 
                                                      stat="identity", position="dodge", col="white") + 
  geom_hline(yintercept=c(0.66, 0.33), linetype="dashed")

ggplot(res, aes(treatment, prop)) + geom_boxplot(aes(col=traj)) + 
  geom_hline(yintercept=c(0.66, 0.33), linetype="dashed") +
  theme_bw()


ggplot(res, aes(treatment, prop)) + geom_boxplot(aes(col=treatment)) + 
  geom_hline(yintercept=c(0.66, 0.33), linetype="dashed") +
  theme_bw()
```


